---
title: "Bird survey detection probability"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(ggplot2)
library(readxl)
library(stringr)
library(tidyr)

library(runjags)


# Default black and white theme for graphs
theme_set( theme_bw() )

    
# convenience function to load contents of an RData file
# into a named object
load_from <- function(.path) {
  load(.path)
  objs <- ls(all.names = FALSE)
  get(objs[1], inherits = FALSE)
}


# Set to TRUE to force re-fitting of models
REFIT_MODEL <- FALSE

```

### Contents

The following R code was used to estimate and graph species and guild detection probabilities in Franklin et al. (in review). It is based on the original version written by Michael Bedward, University of Wollongong.


### Survey data

```{r}

path <- here("all_survey_recs.csv")
DAT.survey <- read.csv(path, stringsAsFactors = FALSE)


# Number of replicates per site under each survey method
Replicates <- DAT.survey %>%
  group_by(site, survey) %>%
  summarize(nreps = max(rep)) %>%
  ungroup() %>%
  tidyr::spread(survey, nreps)


# Species codes, names and info
path <- here("species_lookup.csv")
SpeciesLookup <- read.csv(path, stringsAsFactors = FALSE)

```


### Model code

```{r}

model.code <- "model {
  for (site in 1:nsites) {
    for (sp in 1:nspp) {
      # Model the observations
      for (m in 1:nmethods) {
        num.obs[site,sp,m] ~ dbin(present[site,sp] * p.detect[sp,m], nreps[site,m])
      }

      # Model the true (unknown) presences
      present[site,sp] ~ dbern(p.present[site,sp])
      logit(p.present[site,sp]) <- alpha.present.site[site,sp] + beta.site[site]
      alpha.present.site[site,sp] ~ dnorm(alpha.present[sp], 1)
    }
  }
  
  for (sp in 1:nspp) {
    for (m in 1:nmethods) {
      logit(p.detect[sp,m]) <- alpha.detect.m[sp,m]
      alpha.detect.m[sp,m] ~ dnorm(alpha.detect[sp], 1)
    }
    
    alpha.present[sp] ~ dnorm(0, tau.present)
    alpha.detect[sp] ~ dnorm(0, tau.detect)
  }

  for (site in 1:nsites) {
    beta.site[site] ~ dnorm(0, 1)
  }

  sd.present ~ dexp(1)
  tau.present <- pow(sd.present, -2)

  sd.detect ~ dexp(1)
  tau.detect <- pow(sd.detect, -2)
}"

```


### Assemble data and compile model

```{r}

dat.counts <- DAT.survey %>%
  left_join(SpeciesLookup %>% select(spnum, speciescode), by = "speciescode") %>%
  group_by(site, spnum, survey) %>%
  summarize(nobs = sum(observed)) %>%
  ungroup() %>%
  tidyr::spread(survey, nobs)

# function to format data as a site x sp matrix for each survey method
make_matrix <- function(survey.method) {
  dat.counts %>%
    select(site, spnum, nobs = survey.method) %>%
    tidyr::spread(spnum, nobs) %>%
    select(-site) %>%
    as.matrix()
}

# combine the two matrices into a 3D array for the model
ar.counts <- abind::abind(make_matrix("acoustic"), make_matrix("standard"), along = 3)


model.data <- list(
  nsites = 10,
  nspp = 48,
  nmethods = 2,
  nreps = as.matrix(Replicates[, c("acoustic", "standard")]),
  num.obs = ar.counts
)


ModelPath <- here("model.RData")

if (REFIT_MODEL || !file.exists(ModelPath)) {
  model <- run.jags(
    model = model.code, 
    monitor = c("alpha.detect.m", "alpha.present", "beta.site"), 
    data = model.data, 
    
    burnin = 4000,
    sample = 2000,
    n.chains = 4,
    thin = 20,
    
    method = "parallel",
    
    inits = function() {
      list(present = matrix(1, nrow = 10, ncol = 48))
    },
    
    modules = "glm")
  
  save(model, file = ModelPath)
  
} else {
  # load previously fitted model
  model <- load_from(ModelPath)
}

```


### Check model convergence

```{r}

x <- summary(model)

cat("Min effective samples:", min(x[, "SSeff"]), "\n")
max.gelman <- cat("Max Gelman-Rubin value:", max(x[, "psrf"]), "\n")

```


### Summarize site effects for inspection

In the model, we allow for general site effects such as, for example, a habitat state that increases or reduces the probability of occupancy across all species, by including a term for site.

```{r}

samples <- do.call(rbind, model$mcmc)

ii <- str_detect(colnames(samples), "beta.site")

dat.site <- samples[, ii] %>%
  as.data.frame() %>%
  
  mutate(iter = row_number()) %>%
  
  tidyr::gather(param, value, -iter) %>%
  
  mutate(site = as.integer(str_extract(param, "\\d+"))) %>%
  
  select(iter, site, value)


dat.gg <- dat.site %>%
  group_by(site) %>%
  
  summarize(lwr90 = quantile(value, 0.05),
            lwr50 = quantile(value, 0.25),
            mid = median(value),
            upr50 = quantile(value, 0.75),
            upr90 = quantile(value, 0.95)) %>%
  
  ungroup()


ggplot(data = dat.gg, aes(x = site)) +
  geom_linerange(aes(ymin = lwr90, ymax = upr90),
                 size = 1) +
  
  geom_linerange(aes(ymin = lwr50, ymax = upr50),
                 size = 2.5) +
  
  geom_point(aes(y = mid), size = 4) +
  
  scale_x_continuous(breaks = 1:max(dat.gg$site)) +
  
  labs(y = "Site effect", x = "Site number") +
  
  coord_flip() +
  
  theme(panel.grid.minor = element_blank())

```


### Graph probability of detection

```{r fig.width=6, fig.height=8}

ii <- str_detect(colnames(samples), "alpha.detect")

dat.detect <- samples[, ii] %>%
  as.data.frame() %>%
  
  mutate(iter = row_number()) %>%
  
  tidyr::gather(param, value, -iter) %>%
  
  mutate(inds = str_extract(param, "\\d+,\\d")) %>%
  
  tidyr::separate(inds, into = c("spnum", "survey"), convert = TRUE) %>%
  
  select(iter, spnum, survey, value) %>%
  
  mutate(survey = c("acoustic", "standard")[survey],
         prob = plogis(value))
       

dat.gg <- dat.detect %>%
  group_by(spnum, survey) %>%
  
  summarize(lwr90 = quantile(prob, 0.05),
            lwr50 = quantile(prob, 0.25),
            mid = median(prob),
            upr50 = quantile(prob, 0.75),
            upr90 = quantile(prob, 0.95)) %>%
  
  ungroup() %>%
  
  left_join(SpeciesLookup %>% select(spnum, speciescode), by = "spnum")


Levels <- dat.gg %>%
  filter(survey == "acoustic") %>%
  arrange(mid) %>%
  select(speciescode)


dat.gg$speciescode <- factor(dat.gg$speciescode, levels = Levels$speciescode)

dat.gg.mid <- dat.gg %>%
  select(speciescode, survey, mid) %>%
  tidyr::spread(survey, mid)


  
ggplot(data = dat.gg,  aes(x = speciescode)) +
  geom_linerange(data = dat.gg.mid, 
                 aes(ymin = acoustic, ymax = standard),
                 size = 3, colour = "grey90") +

  geom_linerange(aes(ymin = lwr90, ymax = upr90, colour = survey),
                 position = position_dodge(width = 0.2),
                 size = 1) + 
  
  geom_linerange(aes(ymin = lwr50, ymax = upr50, colour = survey),
                 position = position_dodge(width = 0.2),
                 size = 2.5) + 
 
  coord_flip() +
  
  scale_color_manual(values=c("standard" = "goldenrod1", "acoustic" = "royalblue")) + 
  
  xlab("Species code") + ylab("Probability of detection") + 
  
  theme(legend.position = "none", text=element_text(size=9))

```

### Probability that acoustic survey had better detection rates

Here we calculate the probability that acoustic sampling gives better detection rates for each species

```{r}

 dat <- dat.detect %>%
  select(iter, spnum, survey, prob) %>%
  tidyr::spread(survey, prob) %>%
  
  mutate(diff = acoustic - standard) %>%
  
  group_by(spnum) %>%
  summarize(p.acoustic.better = mean(diff > 0)) %>%
  
  left_join(SpeciesLookup %>% select(spnum, speciescode), by = "spnum") %>%
  
  select(speciescode, p.acoustic.better)

knitr::kable(dat, digits = 2)

```

### Summarize probability of detection by species groupings

Here we query the posterior distribution of the existing, ungrouped species model, and summarize probability of detection for each foraging stratum guild. This involves aggregating the posterior samples according to guild membership. This is somewhat approximate because the guilds differ in number of species. We indicate this in the graph labels.

```{r fig.width=2.9, fig.height=2.5}

dat.n <- SpeciesLookup %>%
  group_by(foragingstratum) %>%
  summarize(count = n())

dat.gg <- dat.detect %>%
  left_join(SpeciesLookup, by = "spnum") %>%
  
  group_by(foragingstratum, survey) %>%
  
  summarize(lwr90 = quantile(prob, 0.05),
            lwr50 = quantile(prob, 0.25),
            mid = median(prob),
            upr50 = quantile(prob, 0.75),
            upr90 = quantile(prob, 0.95)) %>%
  
  ungroup() %>%
  
  left_join(dat.n, by = "foragingstratum") %>%
  mutate(label = paste0(foragingstratum, " (", count, ")"))


dodgew <- 0.5

abrev <- c("A/G (2)", "All (29)", "C/S (8)", "G/U (9)")

ggplot(data = dat.gg, aes(x = label)) +
  geom_linerange(aes(ymin = lwr90, ymax = upr90, colour = survey),
                 position = position_dodge(width = dodgew),
                 size = 1) +
  
  geom_linerange(aes(ymin = lwr50, ymax = upr50, colour = survey),
                 position = position_dodge(width = dodgew),
                 size = 2.5) +
  
  geom_point(aes(y = mid, colour = survey),
                 position = position_dodge(width = dodgew),
                 size = 4) +
  
  
  labs(x = "Foraging stratum guilds", y = "Probability of detection") +
  
  scale_color_manual(values=c("standard" = "goldenrod1", "acoustic" = "royalblue")) + 
  
  theme(legend.position = "none", text=element_text(size=9)) +
  
  coord_flip(ylim = c(0, 1)) +
  
  scale_x_discrete(labels= abrev)

```











